QUESTION HERE????
========================================================

## NOAA Storm Database Analysis
This report analyzes the US National Oceanic and Atmospheric Administration's (NOAA) Storm database which contains characteristics of major US storms and weather events.  This data includes location, fatalities, injuries and property damage.

Data was downloaded as part of the Coursera Reproducible Research project from https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2 on May 9, 2014. 

```{r loadData, echo=TRUE}
# libraries and load data
library(ggplot2)


datafile <- "repdata-data-StormData.csv.bz2"
if (! file.exists(datafile)) { # download it if it does not exist
        url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
        download.file(url, datafile, mode = "wb")
}
# We read the first few rows and then reset the file to only read the columns we care about
# This is just to speed up the lenghty read and to reduce the required memory footprint
data <- read.table(bzfile(datafile), sep=",", header = TRUE, stringsAsFactors=FALSE, nrows=10)

# Now read all the data that we care about
# Originally we had all these fields:
#  "STATE__"    "BGN_DATE"   "BGN_TIME"   "TIME_ZONE"  "COUNTY"     "COUNTYNAME" "STATE"     
#  "EVTYPE"     "BGN_RANGE"  "BGN_AZI"    "BGN_LOCATI" "END_DATE"   "END_TIME"   "COUNTY_END"
#  "COUNTYENDN" "END_RANGE"  "END_AZI"    "END_LOCATI" "LENGTH"     "WIDTH"      "F"         
#  "MAG"        "FATALITIES" "INJURIES"   "PROPDMG"    "PROPDMGEXP" "CROPDMG"    "CROPDMGEXP"
#  "WFO"        "STATEOFFIC" "ZONENAMES"  "LATITUDE"   "LONGITUDE"  "LATITUDE_E" "LONGITUDE_"
#  "REMARKS"    "REFNUM"
# Now we just set the columns we need for our analysis to save space and time
neededCharCols <- c("STATE", "BGN_DATE", "EVTYPE", "F", "PROPDMGEXP")
neededNumCols <- c("LENGTH", "WIDTH", "MAG", "FATALITIES","INJURIES", "PROPDMG", "LATITUDE", "LONGITUDE")
# Set the columns we want, others set to NULL to ignore
classes <- rep("NULL", length(names(data)))
classes[names(data) %in% neededCharCols] <- "factor" # set just the cols we want as char factors
classes[names(data) %in% neededNumCols] <- "numeric" # set just the cols we want as numeric
data <- read.table(bzfile(datafile), sep=",", header = TRUE, colClasses = classes, quote ='"')
rm(classes) # just a cleanup

# Convert the begining date and time to a native format for processing
data$BGN_DATE <- strptime(data$BGN_DATE, format = "%m/%d/%Y %H:%M:%s")

# Basic summary stats
sapply(data, class)
summary(data)
sum(is.na(data))
```




```{r fig.width=7, fig.height=6}

```

